Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds

### 2023-11-30 manuscript/manuscript.qmd
updates
added in fig and table captions, more org.

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  tidy = FALSE,
  message = FALSE,
  fig.height = 4.8,
  fig.width = 6,
  dpi = 300
)
```

@@

Field samples were collected from eight ponds in Rhode Island during October and November 2021. Two liter surface samples were collected in triplicate (six total liters) by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment. If a surface scum was present, the scum was gently brushed aside to avoid being collected in the sample bottles. Samples were collected in acid-washed 1 L amber bottles and placed in a cooler with ice. At the same location as sample collection, measurements of chlorophyll and phycocyanin were made using the AlgaeTorch and the FluoroSense.

@@

## Statistical Analysis

Make some comment about why no 1:1 line, using regression and r2 to compare sensors.

@@

Need summary of sites and data

![Fig 1. Map of Field Sites](images/fig1_map.jpg)

@@

Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond). Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (Figure 2).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.

![Figure 2. Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg)

@@

Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond). The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (Figure 3). Phycocyanin measurements for the Fluorosense (r2=0.96) and the Cyanofluor (r2=0.78) were also strongly related with extracted phycocyanin.

![Figure 3. Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg)

@@

Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(Figure 4).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(Figure 5). Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (Table 1).

![Figure 4. Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){width="2250"}

![Figure 5. Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg)

Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (r2=0.92) followed by the algaetorch (r2=0.78) and the Phycoprobe (r2=0.59). The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (Figure 6). Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(Figure 7).

![Figure 6. Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig6_chl_vs_cells.jpg)

![Figure 7. Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig7_phyco_vs_cells.jpg)

Table 1. Waterbody, Cyano Genus, Relative Biovolume Total Biovolume

```{r show_table1, results='asis'}
knitr::kable(table1)
```

@@

## Acknowledgements

@@

## References

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  tidy = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.height = 4.8,
  fig.width = 6,
  dpi = 300
)
```

<!-- ADD IN CODE SO IT RE-RUNS ON RENDER -->

@@

Field samples were collected from eight ponds in Rhode Island during October and November 2021 (@fig-map). Two liter surface samples were collected in triplicate (six total liters) by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment. If a surface scum was present, the scum was gently brushed aside to avoid being collected in the sample bottles. Samples were collected in acid-washed 1 L amber bottles and placed in a cooler with ice. At the same location as sample collection, measurements of chlorophyll and phycocyanin were made using the AlgaeTorch and the FluoroSense.

![Field sampling locations](images/fig1_map.jpg){#fig-map width=100%}

@@

## Data Analysis

To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots to assess the fit and compare the different sensors. We derived the coefficient of determination from the regressions as a measure of fit and qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts. We do not compare slopes as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.

@@

Need summary of sites and dataFig 1. Map of Field Sites

@@

Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond). Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.

![Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg){#fig-chlascatter width=80%}

@@

Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond). The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (@fig-phycoscatter). Phycocyanin measurements for the Fluorosense (r2=0.96) and the Cyanofluor (r2=0.78) were also strongly related with extracted phycocyanin.

![Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width=80%}

@@

Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-relbv).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(@fig-bv). Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (@tbl-supp).

![Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){#fig-relbv width=80%}

![Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg){#fig-bv width=80%}

Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (r2=0.92) followed by the algaetorch (r2=0.78) and the Phycoprobe (r2=0.59). The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount). Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(@fig-phycocount).

![Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig6_chl_vs_cells.jpg){#fig-chlcount width=80%}

![Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig7_phyco_vs_cells.jpg){#fig-phycocount width=80%}

@@

# Acknowledgements

@@

# Supplement

```{r show_table1, results='asis', echo=FALSE}
#| label: tbl-supp
#| tbl-cap: "Waterbody, Cyano Genus, Relative Biovolume Total Biovolume"

knitr::kable(table1)
```

# References
